---
title: 'Impact of Severe Weather Events: Fatalities, Injuries, and Economic Consequences'
output: word_document
---

###Synopsis 
that describes and summarizes the data analysis

### Data Processing
that describes how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.
```{r processing_code, eval=FALSE}

runObs = 400000 #limit reading datafile

storms <- read.csv("repdata-data-StormData.csv.bz2")#, nrows = runObs)
storms2 <- storms[c("REFNUM", "EVTYPE", "INJURIES", "FATALITIES", "PROPDMG", "CROPDMG")]

#classify EVTYPE into smaller EVGROUPS
library(dplyr)
storms3 <- mutate(storms2, HUMANCOST = INJURIES + FATALITIES, ECONCOST = PROPDMG + CROPDMG, EVGROUP = "Unclassified")

storms3 <- storms3[c("REFNUM", "EVTYPE", "EVGROUP", "HUMANCOST", "ECONCOST")] #FOR PART I
EVGROUP_col = 3
HUMANCOST_col = 4
ECONCOST_col = 5

for(i in 1:nrow(storms3)) {
  
        EVENT_i <- tolower(storms3[i,"EVTYPE"])

        #classify as MARINE
        if(grepl("hurricane", EVENT_i) | grepl("current", EVENT_i) | grepl("tide", EVENT_i) 
           | grepl("sea", EVENT_i) | grepl("marine", EVENT_i) |grepl("surf", EVENT_i) 
           | grepl("coast", EVENT_i) 
         ){
                storms3[i,EVGROUP_col] = "MARINE" #set the EVGROUP column
           }
        
        #classify events as TERRESTRIAL 
        if(grepl("avalanche", EVENT_i) | grepl("fire", EVENT_i) | grepl("mud", EVENT_i) 
           | grepl("pack", EVENT_i) | grepl("flood", EVENT_i)
         ){
                storms3[i,EVGROUP_col] = "TERRESTRIAL" #set the EVGROUP column
           }
        
        #classify as WIND, if applicable
        if(grepl("nado", EVENT_i) | grepl("wind", EVENT_i) | grepl("turbulance", EVENT_i) 
           | grepl("funnel", EVENT_i) | grepl("storm", EVENT_i) |grepl("burst", EVENT_i) 
           | grepl("dust", EVENT_i) | grepl("spout", EVENT_i) |grepl("cloud", EVENT_i) 
         ){
                storms3[i,EVGROUP_col] = "WIND" #set the EVGROUP column
           }
        

        #classify as PRECIPITATION, if applicable
        if(grepl("hail", EVENT_i) | grepl("rain", EVENT_i) | grepl("snow", EVENT_i) 
           | grepl("freez", EVENT_i) | grepl("flood", EVENT_i) |grepl("thunder", EVENT_i) 
           | grepl("preci", EVENT_i) | grepl("wet", EVENT_i) |grepl("lightening", EVENT_i) 
           | grepl("blizzard", EVENT_i) | grepl("sleet", EVENT_i) |grepl("urban", EVENT_i) 
         ){
                storms3[i,EVGROUP_col] = "PRECIPITATION" #set the EVGROUP column
           }
        
        #classify as TEMPERATURE, if applicable
        if(grepl("cold", EVENT_i) | grepl("hot", EVENT_i) | grepl("cool", EVENT_i) 
           | grepl("warm", EVENT_i) | grepl("heat", EVENT_i) | grepl("frost", EVENT_i) 
           | grepl("chill", EVENT_i) | grepl("winter", EVENT_i) | grepl("ice", EVENT_i) 
                | grepl("glaze", EVENT_i)){
                storms3[i,EVGROUP_col] = "TEMPERATURE" #set the EVGROUP column
           }
        
        
        } #end of FOR loop


table(storms3$EVGROUP)
saveRDS(storms3, "storms3_400K.RDS")

```

### Q1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
```{r Human_Cost}
humancost <- tapply(storms3$HUMANCOST, storms3$EVGROUP, "sum")
humancost
qplot(names(humancost), humancost, size = humancost, main = "Human Cost of Weather Events", xlab = "Event Type Group", ylab = "Human Cost: Death and Injury")

```

### Q2 Across the United States, which types of events have the greatest economic consequences?
```{r Economic_Cost}

econcost <- tapply(storms3$ECONCOST, storms3$EVGROUP, "sum")
econcost
qplot(names(econcost), humancost, size = econcost, main = "Economic Cost of Weather Events", xlab = "Event Type Group", ylab = "Economic Costs: Crop and Property Damage")

```
